Syntax and Semantics of Cedille

### 2020-12-22 docs/semantics/paper.tex
docs: semantics: syntactic kind preservation for types under CBN normalization

@@

\end{document}

>>>>

We make a concession to syntactic classifier preservation in the case of types
and kinds.
During type inference, types may be reduced using a call-by-name operational
semantics to reveal type constructors.
With the removal of lifting from CDLE, terms cannot compute types and so no
terms need to be reduced during this process.

\begin{theorem}[Syntactic kind preservation]
  \label{thm:syntactic-kind-pres}
  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadsto_{\beta} T'\) then
  \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa'\) such that \(\kappa
  \cong \kappa'\).
\end{theorem}

@@

\section{Proof of Theorem~\ref{thm:syntactic-kind-pres}}
\label{sec:proof-syntactic-kind-pres}

First a few lemmas (easy proofs omitted):
\begin{lemma}
  \label{lem:inference-validity}
  \ \\
  \begin{itemize}
  \item If \(\Gamma \vdash T \tpsynth \kappa\) then \(\Gamma \vdash \kappa\)

  \item If \(\Gamma \vdash t \ T\) then \(\Gamma \vdash T \tpsynth
    \kappa\) for some \(\kappa\)
  \end{itemize}
\end{lemma}

\begin{lemma}
  \label{lem:subst-kind}
  \ \\
  \begin{itemize}
  \item If \(\Gamma,x:T \vdash T \tpsynth \kappa\) and \(\Gamma \vdash t
    \tpcheck T\) then \(\Gamma \vdash [t/x]T \tpsynth [t/x]\kappa\)

  \item If \(\Gamma,X:\kappa_2 \vdash T_1 \tpsynth \kappa_1\) and \(\Gamma
    \vdash T_2 \tpsynth \kappa_2\) then \(\Gamma \vdash [T_2/X]T_1 \tpsynth [T_2/X]\kappa_1\)
  \end{itemize}
\end{lemma}

\begin{lemma}
  \label{lem:tpcheck-conv}
  If \(\Gamma \vdash t \tpcheck T_1\) where \(\Gamma \vdash T_1 \tpsynth
  \kappa_1\), and we have \(\Gamma \vdash T_2 \tpsynth \kappa_2\) with \(T_1
  \cong T_2\) and \(\kappa_1 \cong \kappa_2\), then \(\Gamma \vdash t \tpcheck T_2\).
\end{lemma}

\begin{proof}[Proof (of Theorem~\ref{thm:syntactic-kind-pres})]
  We may rule out the cases where \(T\) is a variable or formed by a type
  constructor, as this would contradict the assumption that \(T \leadsto_{\beta}
  T'\) for some \(T'\) (recall that \(\leadsto_\beta\) is call-by-name reduction
  for types).

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa_1}{\Gamma\vdash T \tpsynth
      \abs{\Pi}{x}{T_1}{\kappa_1} & \Gamma\vdash t \tpcheck T_1}
  \]

  There are two subcases to consider for the derivation of \(T\ t \leadsto_\beta
  T'\). In the first case, we have \(T \leadsto_\beta T''\) for some \(T''\) (so
  \(T' = T''\ t\)). By the IH, we have \(\Gamma \vdash T'' \tpsynth
  \abs{\Pi}{x}{T_1'}{\kappa_1'}\) for some \(T_1' \cong T_1\) and \(\kappa_1'
  \cong \kappa_1\) (we have by assumption the kind of \(T''\) must be
  convertible with \(\abs{\Pi}{x}{T_1}{\kappa_1}\)). By
  Lemma~\ref{lem:tpcheck-conv}, \(\Gamma \vdash t \tpcheck T_1'\). By the rule,
  \(\Gamma \vdash T''\ t \tpsynth [t/x]\kappa_1'\), and the synthesized kind is
  clearly convertible with \([t/x]\kappa_1\).

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{x}{T_1}{T''})\ t\) and our assumed reduction is to
  \([t/x]T''\). By inversion of the kinding derivation, we have \(\Gamma,x:T_1
  \vdash T'' \tpsynth \kappa_1\). By Lemma~\ref{lem:subst-kind}, we have
  \(\Gamma \vdash [t/x]T'' \tpsynth [t/x]\kappa_1\).

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1
      \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth
      \kappa_2' & \kappa_2 \cong \kappa_2'}
  \]

  There are two subcases to consider for the derivation of \(T_1 \cdot T_2
  \leadsto_\beta T'\). In the first subcase, we have \(T_1 \leadsto_\beta
  T_1''\) for some \(T_1''\) (so \(T' = T_1'' \cdot T_2\)). By the IH, we have
  \(\Gamma \vdash T_1'' \tpsynth \abs{\Pi}{X}{\kappa_2''}{\kappa_1''}\) for some
  \(\kappa_2'' \cong \kappa_2\) and \(\kappa_1'' \cong \kappa_1\) (we have by
  assumption the kind of \(T_1''\) must be convertible with
  \(\abs{\Pi}{X}{\kappa_2}{\kappa_1}\)). By transitivity we can conclude
  \(\kappa_2'' \cong \kappa_2\). By the rule, we have \(\Gamma \vdash T_1''
  \cdot T_2 \tpsynth [T_2/X]\kappa_1''\), and the synthesized kind is clearly
  convertible with \([T_2/X]\kappa_1\).

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
  to \([T_2/X]T_1''\). By inversion of the kinding derivation, we have
  \(\Gamma,X:\kappa_2 \vdash T_1'' \tpsynth \kappa_1\). By
  Lemma~\ref{lem:subst-kind}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
  [T_2/X]\kappa_1\).

\end{proof}
\end{document}

% LocalWords:  nondeterministically Kleene's realizability realizer kinded CDLE
% LocalWords:  Cedille subcases subcase kinding
